---
title: "analysis-2.rmd"
output: html_document
date: "2025-10-28"
---

## Analysis 2 for GRI

The preprocess-1.ipynb and analysis-1.ipynb preceed this analysis.
The previous files are used for the data and metadata generation in python. 
In this file, some inital steps for the analysis are carried out in R, since traditionally survival analysis is done in R, the resources are richer. 


Load the data generated in python

```{r}
library(dplyr)
library(tidyr)
```

```{r}

df = read.csv("data/GRI_preprocess-2.csv", na.strings = "")

rownames(df) = df$TMA_order

df = df[, !(names(df) %in% c("TMA_order"))]

df = df[, !(names(df)) %in% c("X")]

```

```{r}
binary <- c('Emergency_Surgery', 'Sex', 'Tumour_Budding',
       'Tumour_Perforation', 'Margin_Invasion', 'Venous_Invasion',
       'Peritoneal_Invasion', 'IL6R_AK_186_64', 'ZEB1_Nuclear',
       'Twist_Cytoplasmic', 'Twist_Nuclear', 'Snail_Cytoplasmic',
       'Snail_Nuclear', 'SOX9_Cytoplasmic', 'SOX9_Nuclear', 'ZEB1_Cytoplasmic',
       'BudCohort_ZEB1_Cytoplasmic_A', 'BudCohort_ZEB1_Nuclear',
       'IL17RA_Presence', 'BRAF', 'KRAS', 'NRAS')

```

```{r}
cat <- c('MMR_Status', 'Tumour_Location', 'T_Stage', 'N_Stage',
       'Histological_Grade', 'Elastica_Venous_Invasion')
```

```{r}
# transform to factors
df[binary] <- lapply(df[binary], as.factor)
df[cat] <- lapply(df[cat], as.factor)
```



### Descriptive statistics

```{r}
library(gtsummary)

summary_table <- df %>% 
tbl_summary(
  by = Cmprsk_Status,
  statistic = list(all_continuous() ~ "{mean} ({sd})",
                   all_categorical() ~ "{n} ({p}%)"
))

# latex
latex_code <- summary_table %>%
  as_kable(format = "latex", booktabs = TRUE, caption = "Summary Table by Overall Survival Status", label = "tab:summary")


```

```{r}
summary_table
```


## Kaplan Meier curves

The competing risk is going to be censored, we are overestimating the event of interest incidence by using Kaplan-Meier.

```{r}
library(survival)
library(survminer)
library(ggplot2)
```

```{r, fig.height=5, fig.weight=7}

SurvObj <- with(df, Surv(time = Disease_Free_Survival_Months, event = Cmprsk_Status == 1))

fit_sex <- survfit(SurvObj ~ Sex, data = df)

summary(fit_sex)

survdiff(SurvObj ~ Sex, data = df)

ggsurvplot(fit_sex, data = df,
           pval = TRUE,
           conf.int = TRUE,
           risk.table = TRUE,
           ggtheme = theme_minimal(),
           title = "Kaplan–Meier by Sex")
```

Do the same for all variables

We are clearly considering competing risks as censored and performing normal survival analysis:

```{r}
table(df$Histological_Grade, useNA = 'ifany')
```


```{r, fig.height=5, fig.weight=7}

# time_col   <- "Disease_Free_Survival_Months"
# status_col <- "Cmprsk_Status"
# 
# cat_vars <- names(df)[sapply(df, \(x) is.factor(x) || is.character(x))]
# cat_vars <- setdiff(cat_vars, c(time_col, status_col))
# 
# for (v in cat_vars) {
#   dat <- df[, c(time_col, status_col, v)]
# 
#   # make explicit columns so no symbols are looked up
#   dat$time  <- dat[[time_col]]
#   dat$event <- dat[[status_col]] == 1
#   dat$grp   <- droplevels(dat[[v]])
# 
#   fit <- survfit(Surv(time, event) ~ grp, data = dat)
# 
#   pl <- ggsurvplot(
#     fit, data = dat,
#     pval = TRUE, conf.int = TRUE, risk.table = TRUE,
#     ggtheme = theme_minimal(), title = paste("KM by", v), legend.title = v
#   )
# 
#   print(pl)
# }

```

```{r}
# for (v in cat_vars) {
#   dat <- df[, c(time_col, status_col, v)]
#   dat$time  <- dat[[time_col]]
#   dat$status <- dat[[status_col]]
#   dat$grp   <- droplevels(dat[[v]])
# 
#   ci <- cmprsk::cuminc(
#     ftime   = dat$time,
#     fstatus = dat$status,
#     group   = dat$grp
#   )
# 
#   plot(ci, wh = 1, main = paste0(v, " Cumulative incidence (cause 1)"), xlab = "Time (months)",
#      ylab = "Probability", lty=1:length(ci), color=1, lwd=par('lwd'),)
# 
# }
```


```{r  fig.height=5, fig.weight=7}
library(tidycmprsk)
library(ggplot2)
library(ggsurvfit)
library(cmprsk)

time_col   <- "Disease_Free_Survival_Months"
status_col <- "Cmprsk_Status"

df2 <- df %>%
  mutate(
    !!status_col := factor(
      .data[[status_col]],
      levels = c(0, 1, 2), 
      labels = c("censor", "death from cancer", "death other causes")
    )
  ) %>%
  filter(is.finite(.data[[time_col]]), !is.na(.data[[status_col]]))

outcomes_present <- setdiff(levels(df2[[status_col]]), "censor")
cause_labels <- c("death from cancer", "death other causes")

grp_var <- "Sex"
df2[[grp_var]] <- factor(df2[[grp_var]])

f <- as.formula(paste0("Surv(", time_col, ", ", status_col, ") ~ ", grp_var))

ci_obj <- tidycmprsk::cuminc(f, data = df2)

pvals <- ci_obj$cmprsk$Tests[, "pv"] 
subtitle_txt <- paste(paste0(cause_labels, ": p=", signif(pvals, 3)), collapse = "   ")

p <- ggsurvfit::ggcuminc(ci_obj, outcome = c("death from cancer", "death other causes")) +
  add_confidence_interval() +
  add_risktable() +
  scale_ggsurvfit() +
  labs(
    title = paste("Cumulative incidence by", grp_var),
    subtitle = subtitle_txt,
    x = "Time (months)",
    y = "Cumulative incidence"
  )

print(p)

```

Let's perform this with all categoricals: 


```{r fig.height=5, fig.weight=7}
time_col   <- "Disease_Free_Survival_Months"
status_col <- "Cmprsk_Status"

df2 <- df %>%
  mutate(
    !!status_col := factor(
      .data[[status_col]],
      levels = c(0, 1, 2), 
      labels = c("censor", "death CRC/recurrence", "death other causes")
    )
  ) %>%
  filter(is.finite(.data[[time_col]]), !is.na(.data[[status_col]]))

outcomes_present <- setdiff(levels(df2[[status_col]]), "censor")
cause_labels <- c("death CRC/recurrence", "death other causes")

cat_vars <- names(df2)[vapply(df2, function(x) is.factor(x) || is.character(x) || is.logical(x), TRUE)]
cat_vars <- setdiff(cat_vars, c(time_col, status_col))


for (v in cat_vars) {

  f <- as.formula(paste0("Surv(", time_col, ", ", status_col, ") ~ ", v))
  
  ci_obj <- tidycmprsk::cuminc(f, data = df2)
  
  pvals <- ci_obj$cmprsk$Tests[, "pv"] 
  subtitle_txt <- paste(paste0(cause_labels, ": p=", signif(pvals, 3)), 
                        collapse = "   ")
  
  p <- ggcuminc(ci_obj, outcome = "death CRC/recurrence")+
    add_confidence_interval() +
    add_risktable() +
    scale_ggsurvfit() +
    labs(
      title = paste("Cumulative incidence by", v),
      subtitle = subtitle_txt,
      x = "Time (months)",
      y = "Cumulative incidence"
    )

  print(p)
}
```

### bring the coded variables... ? 


### Simple train/test split to test addition of covariates:


```{r}
library(recipes)
library(rsample)
library(survival)
library(cmprsk)
library(broom)
```

```{r}
set.seed(123)

rownames(df) <- df$TMA_Order

df <- df[, !(names(df) %in% c("TMA_Order"))]

time_col   <- "Disease_Free_Survival_Months"
status_col <- "Cmprsk_Status"

predictors <- setdiff(names(df), c(time_col, status_col))

predictors
```


```{r}
set.seed(42)

split <- initial_split(df, prop = 0.7, strata = all_of(status_col))
train <- training(split)
test  <- testing(split)

# imputation to be applied to train

rx <- recipe(reformulate(predictors, response = time_col), data = train) %>%
  # the outcome as time
  update_role(all_of(time_col), new_role = "outcome") %>%
  # the predictor is all the rest
  update_role(all_of(setdiff(predictors, time_col)), new_role = "predictor") %>%
  # median and mode
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  # normalise the continuous
  step_normalize(all_numeric_predictors(), -all_outcomes())
  
# imputation and scaling only from the training
rx_prep <- prep(rx, training = train, verbose = FALSE)

X_train <- bake(rx_prep, new_data = train) %>% select(-all_of(time_col))
X_test  <- bake(rx_prep, new_data = test)  %>% select(-all_of(time_col))

ftime_train   <- train[[time_col]]
fstatus_train <- train[[status_col]]

ftime_test    <- test[[time_col]]
fstatus_test  <- test[[status_col]]

```


Avoid collinearity:

```{r}
causes <- sort(setdiff(unique(fstatus_train), 0))

form_all <- as.formula(paste0("~ ", paste(colnames(X_train), collapse = " + ")))
MM <- model.matrix(form_all, data = X_train)[, -1, drop = FALSE]

# avoiding collinearity:
# but how has it been introduced? 
qrX <- qr(MM)
keep <- sort(qrX$pivot[seq_len(qrX$rank)])
predictors_train <- MM[, keep, drop = FALSE]

dif <- setdiff(colnames(MM), colnames(predictors_train))

dif
```

```{r}
# cor(train$Preoperative_CRP, train$CRP,
#   method = "spearman"
# )
# 
# cor(train$Preoperative_Lymphocytes, train$Lymphocytes,
#   method = "spearman"
# )

```
```{r}
plot(train$PTPN12_Tumour, train$PTPN12_Stroma)

cor(train$PTPN12_Tumour, train$PTPN12_Stroma, method = "spearman")
```
```{r}
#plot(log(train$Ki67_Image_Analysis), log(train$Ki67_Tumour))
```
```{r}
#plot(train$CD3FOXP3_Raw, train$CD3_FOXP3_Raw)
```
```{r}
#cor(train$CD3FOXP3_Raw, train$CD3_FOXP3_Raw, use="complete.obs")
```

```{r}
#plot(train$Preoperative_CRP, train$CRP)
```

```{r}
#plot(train$Preoperative_Lymphocytes, train$Lymphocytes)

```


## Missigness

```{r, fig.height=8}
library(naniar)

gg_miss_var(df, show_pct = TRUE)


```

## Correlation

```{r, fig.height=8}
library(reshape2)

num <- df %>%
  select(where(is.numeric))

corr <- stats::cor(num, use = "pairwise.complete.obs", method = "spearman")
melted <- melt(corr)

ggplot(melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Spearman ρ"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation matrix (numeric covariates)", x = NULL, y = NULL)
```

SDMA and Beta Catenines might require selection. 


## Class imbalance

```{r, fig.height=8}
library(forcats)

df_long <- df %>%
  select(where(\(x) is.factor(x) || is.character(x))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  mutate(value = fct_explicit_na(as.factor(value), "NA"))

df_plot <- df_long %>%
  count(variable, value) %>%
  group_by(variable) %>%
  mutate(prop = n / sum(n))

ggplot(df_plot, aes(x = value, y = prop, fill = value)) +
  geom_col() +
  facet_wrap(~ variable, scales = "free_x") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            vjust = -0.3, size = 2.8) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set3", na.value = "grey80") +
  labs(x = NULL, y = "Proportion", title = "Class imbalance across categorical variables") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

```


## Continue analysis

We should keep the preoperatives
```{r}
library(prodlim)
## fine and grey with cmprsk
# fit_fg <- lapply(causes, function(k) {
#   cmprsk::crr(
#     ftime   = ftime_train,
#     fstatus = fstatus_train,
#     cov1    = predictors_train,
#     failcode = k,
#     cencode  = 0
#   )
# })
# names(fit_fg) <- paste0("FG_cause_", causes)

## fine and grey with wrapper from riskregression
train_df <- cbind.data.frame(
  time   = ftime_train,
  status = fstatus_train,
  as.data.frame(X_train)
)

predictors <- setdiff(names(train_df), c("time","status"))

frm <- as.formula(paste0("Hist(time, status) ~ ", paste(predictors, collapse = " + ")))

fit_fg <- lapply(causes, function(k) {
  riskRegression::FGR(frm, data = train_df, cause = k)
})
names(fit_fg) <- paste0("FG_cause_", causes)

```

```{r}
length(predictors)
```


https://pmc.ncbi.nlm.nih.gov/articles/PMC5698744/

```{r, fig.height=8}

fg1 <- fit_fg$FG_cause_1$crrFit
s   <- summary(fg1)
tab <- as.data.frame(s$coef)
colnames(tab) <- make.names(colnames(tab), unique = TRUE)

#beta <- drop(fg1$coef)
#V    <- fg1$var
#se   <- sqrt(diag(V))
#z    <- beta / se
#pval <- 2 * pnorm(abs(z), lower.tail = FALSE)

HR <- data.frame(
  term       = rownames(tab),
  estimate   = tab[["exp.coef."]],
  conf.low   = exp(tab[["coef"]] - 1.96 * tab[["se.coef."]]),
  conf.high  = exp(tab[["coef"]] + 1.96 * tab[["se.coef."]]),
  p.value   = tab[["p.value"]],
  row.names  = NULL,
  check.names = FALSE
)

alpha <- 0.05
HR$sig <- ifelse(is.na(HR$p.value), "NA", ifelse(HR$p.value < alpha, "p<0.05", "ns"))
HR$sig <- factor(HR$sig, levels = c("p<0.05", "ns", "NA"))

ggplot(HR, aes(x = estimate, y = reorder(term, estimate), color = HR$sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), na.rm = TRUE) +
  scale_x_log10() +
  scale_color_manual(
    values = c("p<0.05" = "#D55E00", "ns" = "#555555", "NA" = "#AAAAAA"),
    name = "Significance"
  ) +
  labs(x = "Hazard Ratio (log scale)", y = NULL, title = "Fine–Gray: Cause 1") +
  theme_bw(12) +
  theme(legend.position = "right")
```


```{r}
table(train_df$T_Stage, useNA="ifany")
```

```{r}
table(train_df$Tumour_Perforation, useNA="ifany")
```

```{r}
library(riskRegression)
library(prodlim)

# eval times
times <- quantile(ftime_test, probs = seq(0.1, 0.9, by = 0.05), na.rm = TRUE)
times <- sort(unique(as.numeric(times)))

X_test <- X_test[, colnames(X_train)]

test_df <- cbind.data.frame(
  time   = ftime_test,
  status = fstatus_test,
  as.data.frame(X_test)
)


cif_fg1 <- predictRisk(fit_fg[["FG_cause_1"]], 
                       newdata = test_df, times = times)

#cif_fg1
```


```{r}
sc_fg <- Score(
  object      = list(FG = fit_fg[["FG_cause_1"]]),
  formula     = Hist(time, status) ~ 1,
  data        = test_df,
  times       = times,
  metrics     = "Brier",
  cause       = 1L,
  cens.model  = "km",
  summary     = "ibs",
  split.method = "none"
)
```

```{r}
sc_fg
```
```{r}

tail(sc_fg$Brier$score$IBS, 1)

```


### Additive performance of routine vs research variables.

```{r}
meta <- read.csv("data/GRI_preprocess-1-meta.csv", stringsAsFactors = FALSE)


routine_vars  <- meta %>% filter(Collection == "routine")
research_vars <- meta %>% filter(Collection == "research")
both_vars <- c("Disease_Free_Survival_Months", "Cmprsk_Status", "Lymph_Node_Ratio")


rou <- union(routine_vars$New.Data.Item, both_vars)

rou

```
```{r}
predictors

```


```{r}

res <- union(research_vars$New.Data.Item, both_vars)
res
```

```{r}
both <- intersect(c(union(rou, res)), names(train_df))

both
```

Using the same splits as before:

```{r}
library(prodlim)


train_df <- cbind.data.frame(
  time   = ftime_train,
  status = fstatus_train,
  as.data.frame(X_train)
)

#train_df <- train_df[, !(names(train_df) %in% drop_vars)]


fit_models <- list()
for (nm in c("Routine", "Research", "Both")) {
  predictors <- switch(
    nm,
    "Routine"  = intersect(c(rou), names(train_df)),
    "Research" = intersect(c(res), names(train_df)),
    "Both"     = intersect(c(union(rou, res)), names(train_df))
  )
  
  print(predictors)
  
  frm <- as.formula(paste0("Hist(time, status) ~ ", paste(predictors, 
                                                          collapse = " + ")))
  fit_models[[nm]] <- riskRegression::FGR(frm, data = train_df, cause = 1)

  
}

# test prediction
test_df <- cbind.data.frame(
  time   = ftime_test,
  status = fstatus_test,
  as.data.frame(X_test)
)
#test_df <- test_df[, !(names(test_df) %in% drop_vars)]

# evaluation
times_eval <- c(36, 60) 

scores <- riskRegression::Score(
  object  = fit_models,
  formula = Hist(time, status) ~ 1,
  data    = test_df,
  times   = times_eval,
  metrics = "Brier",
  cause   = 1,
  summary = "ibs"
)

scores$Brier

riskRegression::plotBrier(scores)

```

```{r}
df_brier <- scores$Brier$score

ggplot(df_brier, aes(x = factor(times), y = Brier, color = model, group = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  scale_y_continuous(name = "Brier score ", expand = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(name = "Time (months)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Brier score comparison across models")

```


```{r}
tidy_fg_hr <- function(fg_obj, model_name = "") {
  # accept FGR (wraps cmprsk::crr) or crr directly
  crr <- if (inherits(fg_obj, "FGR")) {
    if (!is.null(fg_obj$crrfit)) fg_obj$crrfit else fg_obj$crrFit
  } else fg_obj
  stopifnot(inherits(crr, "crr"))

  s   <- summary(crr)
  tab <- as.data.frame(s$coef)
  colnames(tab) <- make.names(colnames(tab), unique = TRUE)  # coef, exp.coef., se.coef., z, p (or p.value)

  # robust p-value pick
  pcol <- intersect(c("p","p.value","Pr..>.z."), colnames(tab))
  pvec <- if (length(pcol)) tab[[pcol[1]]] else NA_real_

  out <- data.frame(
    term      = rownames(tab),
    estimate  = tab[["exp.coef."]],
    conf.low  = exp(tab[["coef"]] - 1.96 * tab[["se.coef."]]),
    conf.high = exp(tab[["coef"]] + 1.96 * tab[["se.coef."]]),
    p.value   = pvec,
    model     = model_name,
    row.names = NULL,
    check.names = FALSE
  )
  out$sig <- factor(ifelse(is.na(out$p.value), "NA",
                           ifelse(out$p.value < 0.05, "p<0.05", "ns")),
                    levels = c("p<0.05","ns","NA"))
  out
}
```

```{r}
models3 <- list(
  Routine  = fit_models[["Routine"]],
  Research = fit_models[["Research"]],
  Both     = fit_models[["Both"]]
)
HR_all <- bind_rows(lapply(names(models3), function(nm) tidy_fg_hr(models3[[nm]], nm)))

HR_all <- HR_all %>%
  group_by(model) %>%
  mutate(term_f = factor(term, levels = term[order(estimate)])) %>%
  ungroup()

ggplot(HR_all, aes(x = estimate, y = term_f, color = sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), na.rm = TRUE) +
  scale_x_log10() +
  scale_color_manual(values = c("p<0.05" = "#D55E00", "ns" = "#555555", "NA" = "#AAAAAA"),
                     name = "Significance") +
  labs(x = "Hazard Ratio (log scale)", y = NULL, title = "Fine–Gray (Cause 1): HRs by model") +
  facet_wrap(~ model, scales = "free_y") +
  theme_bw(12) +
  theme(legend.position = "right")
```

