---
title: "analysis-2.rmd"
output: html_document
date: "2025-10-28"
---

## Analysis 2 for GRI

The preprocess-1.ipynb and analysis-1.ipynb proceed this analysis.
The previous files are used for the data and metadata generation in python. 
In this file, some inital steps for the analysis are carried out in R, since traditionally survival analysis is done in R, the resources are richer. 


Load the data generated in python

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}

df = read.csv("data/GRI_preprocess-2.csv", na.strings = "")

rownames(df) = df$TMA_order

df = df[, !(names(df) %in% c("TMA_order"))]

df = df[, !(names(df)) %in% c("X", "Peritoneal_Invasion")]

```

```{r}
colnames(df)
```

```{r}
binary <- c('Emergency_Surgery', 'Sex', 'Tumour_Budding', 'Margin_Invasion','IL6R_AK_186_64', 'ZEB1_Nuclear',
            # 'Peritoneal_Invasion',
       'Twist_Cytoplasmic', 'Twist_Nuclear', 'Snail_Cytoplasmic',
       'Snail_Nuclear', 'SOX9_Cytoplasmic', 'SOX9_Nuclear', 'ZEB1_Cytoplasmic', 'BudCohort_ZEB1_Nuclear',
       'IL17RA_Presence', 'BRAF', 'KRAS', 'NRAS')

```

```{r}
cat <- c('MMR_Status', 'Tumour_Location', 'T_Stage', 'N_Stage',
       'Histological_Grade', 'Elastica_Venous_Invasion')
```

```{r}
# transform to factors
df[binary] <- lapply(df[binary], as.factor)
df[cat] <- lapply(df[cat], as.factor)
```

```{r}
length(colnames(df))
```


### Descriptive statistics

```{r}
library(gtsummary)

summary_table <- df %>% 
tbl_summary(
  by = Outcome,
  statistic = list(all_continuous() ~ "{mean} ({sd})",
                   all_categorical() ~ "{n} ({p}%)"
))

# latex
latex_code <- summary_table %>%
  as_kable(format = "latex", booktabs = TRUE, caption = "Summary Table by Outcome", label = "tab:summary")


```

```{r}
summary_table
```
## Outcomes

```{r}
e <- dim(df[df$Outcome == 0,])[1]/dim(df)[1]*100
f <- dim(df[df$Outcome == 1,])[1]/dim(df)[1]*100
g <- dim(df[df$Outcome == 2,])[1]/dim(df)[1]*100

overall_data <- data.frame(
  Category = rep("Outcomes", 3),
  Outcome = factor(c("Died of CRC/recurrence", "Died of other causes (no recurrence)", 
                     "Alive (no recurrence)"),
                   levels = c("Died of CRC/recurrence", "Died of other causes (no recurrence)",
                              "Alive (no recurrence)")),
  Percentage = c(f, g, e)
)

overall_plot <- ggplot(overall_data, aes(x = Category, y = Percentage, fill = Outcome)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), position = position_stack(vjust = 0.5)) +
  labs(
    #title = "Overall Patient CRC Outcomes After Curative Surgery",
    x = "Category",
    y = "Percentage"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#FB8072", "#80B1D3", "#8DD3C7"))
overall_plot
```
## Recurrence

Load the file with recurrence data, it is the same as the previous with the recurrence information that has been excluded from predictors. 

```{r}
data = read.csv("data/GRI_preprocess-1.csv", na.strings = "")

rownames(data) = data$TMA_order

data = data[, !(names(data) %in% c("TMA_order"))]

data = data[, !(names(data)) %in% c("X")]

```

```{r}
table(data$REC_n0_yes1) / nrow(data) * 100
```

```{r}
library(survival)
library(survRM2)

data$rec_group <- ifelse(!is.na(data$REC_n0_yes1) & data$REC_n0_yes1 == 1, 1, 0)
time   <- data$Cmprsk_time
status <- ifelse(data$UpdateJuly2020_a0_cd1_ncd2 == 1, 1, 0)
group  <- data$rec_group 

# 5 years
tau <- min(60, max(time, na.rm = TRUE))

rm <- rmst2(time = time,
            status = status,
            arm = group,
            tau = tau)

# RMST for non-recurrent group (group==0)
rm$RMST.arm1 
# RMST for recurrent group (group==1)
rm$RMST.arm2 
 # difference, ratio, p-value
rm$unadjusted.result

```



Outcome when died of CRC and recurrence has not yet been grouped. 

```{r}
e <- dim(data[data$UpdateJuly2020_a0_cd1_ncd2 == 0,])[1]/dim(data)[1]*100
f <- dim(data[data$UpdateJuly2020_a0_cd1_ncd2 == 1,])[1]/dim(data)[1]*100
g <- dim(data[data$UpdateJuly2020_a0_cd1_ncd2 == 2,])[1]/dim(data)[1]*100

overall_data <- data.frame(
  Category = rep("Outcomes", 3),
  Outcome = factor(c("Died of CRC", "Died of other causes", "Alive"),
                   levels = c("Died of CRC", "Died of other causes", "Alive")),
  Percentage = c(f, g, e)
)

overall_plot <- ggplot(overall_data, aes(x = Category, y = Percentage, fill = Outcome)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), position = position_stack(vjust = 0.5)) +
  labs(
    #title = "Overall Patient CRC Outcomes After Curative Surgery",
    x = "Category",
    y = "Percentage"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#FB8072", "#80B1D3", "#8DD3C7"))
overall_plot
```
```{r}
df_plot <- data %>%
  mutate(
    Recurrence = case_when(
      REC_n0_yes1 == 1 ~ "Yes",
      TRUE ~ "No"                
    )
  ) %>%
  count(Recurrence) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1))

df_plot$Recurrence <- factor(df_plot$Recurrence, levels = c("No", "Yes"))

cols <- c("No" = "#CCCCCC", "Yes" = "#F4A460")

ggplot(df_plot, aes(x = 2, y = Percentage, fill = Recurrence)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(Percentage, "%")),
            position = position_stack(vjust = 0.5),
            size = 5) +
  scale_fill_manual(values = cols) +
  xlim(0.5, 2.5) +
  theme_void() +
  labs(title = "CRC Patients Progression After Curative Surgery",
       fill = "Recurrence") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    legend.position = "right"
  )
```

```{r}
# percentage of patients that had recurrence and are alive
a <- dim(data[data$REC_n0_yes1 == 1 & data$UpdateJuly2020_a0_cd1_ncd2 == 0, ])[1]/dim(data[data$REC_n0_yes1 == 1, ])[1]*100

b <- dim(data[data$REC_n0_yes1 == 1 & data$UpdateJuly2020_a0_cd1_ncd2 == 1, ])[1]/dim(data[data$REC_n0_yes1 == 1, ])[1]*100

c <- dim(data[data$REC_n0_yes1 == 1 & data$UpdateJuly2020_a0_cd1_ncd2 == 2, ])[1]/dim(data[data$REC_n0_yes1 == 1, ])[1]*100

# Data for recurrence outcomes
recurrence_data <- data.frame(
  Category = rep("Outcomes of patients with recurrence", 3),
  Outcome = factor(c("Died of CRC", "Died of other causes", "Alive"),
                   levels = c( "Died of CRC", "Died of other causes", "Alive")),
  Percentage = c(b, c, a)
)

recurrence_plot <- ggplot(recurrence_data, aes(x = Category, y = Percentage, fill = Outcome)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), position = position_stack(vjust = 0.5)) +
  theme_minimal() +
  scale_fill_manual(values = c("#FB8072", "#80B1D3", "#8DD3C7"))

print(recurrence_plot)
```

```{r}
data2 <- data[, c("REC_n0_yes1", 'Cmprsk_time')]

# Patients with recurrence
recurrence_df <- data2[!is.na(data2$REC_n0_yes) & data2$REC_n0_yes == 1, ]

# Recurrences within each year
months <- c(1/12, 6, 12, 36, 60, 120, 180, 240)
recurrences <- sapply(months, function(x) sum(recurrence_df$Cmprsk_time <= x & recurrence_df$REC_n0_yes == 1))

proportions <- recurrences / nrow(data2) * 100

plot_data <- data.frame(
  Year = c(1/12, 0.5, 1, 3, 5, 10, 15, 20),
  Proportion = proportions
)

ggplot(plot_data, aes(x = Year, y = Proportion)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(#title = "Proportion of Recurrences in GRI over time of study",
       x = "Time (years)",
       y = "Proportion of Recurrences (%)") +
      theme_minimal() +
      scale_x_continuous(breaks = c(1/12, 6/12, 1, 3, 5, 10, 15, 20),
                     labels = c("1m", "", "1y", "3y", "5y","10y", "15y", "20y")) +
      scale_y_continuous(breaks = c(0, 10, 25, 50, 75, 100),
                     limits = c(0, 100), expand = c(0, 0)) +
  geom_vline(xintercept = c(1/12, 0.5, 1, 3, 5), linetype = "dashed", color = "gray") +
  geom_text(data = plot_data[plot_data$Year %in% c(1/12, 0.5, 1, 3, 5, 10, 15, 20), ],
            aes(x = Year, y = Proportion, label = paste(round(Proportion, 2), "%")),
            vjust = -2.5)
```

```{r}
data2 <- data[, c("REC_n0_yes1", "Cmprsk_time", "Recurrence_local1_distant2")]

# Patients with recurrence
recurrence_df <- data2[!is.na(data2$REC_n0_yes) & data2$REC_n0_yes == 1, ]

# Define the recurrence types
recurrence_types <- c(1, 2, 3, 4)
type_labels <- c("Local", "Distant", "Local&Distant", "Unknown")

months <- c(0, 6, 12, 36, 60, 120, 180, 240)
years <- c(0, 0.5, 1, 3, 5, 10, 15, 20)

plot_data_list <- list()

# Loop over each recurrence type
for (i in 1:length(recurrence_types)) {
  type <- recurrence_types[i]
  label <- type_labels[i]
  
  # Filter the data for the current recurrence type
  type_df <- recurrence_df[recurrence_df$Recurrence_local1_distant2 == type, ]
  
  # Calculate the number of recurrences within each year
  recurrences <- sapply(months, function(x) sum(type_df$Cmprsk_time <= x))
  
  # Calculate the proportions
  proportions <- recurrences / nrow(recurrence_df) * 100
  
  plot_data <- data.frame(
    Year = years,
    Proportion = proportions,
    Type = label
  )
  
  # Append to the list
  plot_data_list[[i]] <- plot_data
}

# Combine all plot data into one data frame
combined_plot_data <- do.call(rbind, plot_data_list)

# Plot the cumulative incidence over time 
data2 <- data[, c("REC_n0_yes1", "Cmprsk_time", "Recurrence_local1_distant2")]

# Patients with recurrence
recurrence_df <- data2[!is.na(data2$REC_n0_yes) & data2$REC_n0_yes == 1, ]

# Define the recurrence types
recurrence_types <- c(1, 2, 3, 4)
type_labels <- c("Local", "Distant", "Local&Distant", "Unknown")

# Define the months and years
months <- c(0, 6, 12, 36, 60, 120, 180, 240)
years <- c(0, 0.5, 1, 3, 5, 10, 15, 20)

# Initialize an empty list to store plot data for each type
plot_data_list <- list()

# Loop over each recurrence type
for (i in 1:length(recurrence_types)) {
  type <- recurrence_types[i]
  label <- type_labels[i]
  
  type_df <- recurrence_df[recurrence_df$Recurrence_local1_distant2 == type, ]
  
  recurrences <- sapply(months, function(x) sum(type_df$Cmprsk_time <= x))
  
  proportions <- recurrences / nrow(recurrence_df) * 100
  
  plot_data <- data.frame(
    Year = years,
    Proportion = proportions,
    Type = label
  )
  
  plot_data_list[[i]] <- plot_data
}

combined_plot_data <- do.call(rbind, plot_data_list)

# Plot the cumulative incidence over time using ggplot2
ggplot(combined_plot_data, aes(x = Year, y = Proportion, color = Type, group = Type)) +
  geom_line() +
  geom_point() +
  labs(#title = "Proportion of Recurrences in GRI over time of study",
       x = "Time (years)",
       y = "Proportion of Recurrences (%)",
       color = "Recurrence Type") +
  theme_minimal() +
  scale_x_continuous(breaks = c(1/12, 1, 3, 5, 10, 15, 20),
                     labels = c("1m", "1y", "3y", "5y","10y", "15y", "20y")) +
  scale_y_continuous(breaks = c(0, 10, 25, 50, 75, 100),
                     limits = c(0, 100), expand = c(0, 0)) +
  geom_text(
    data = combined_plot_data[combined_plot_data$Type == "Distant" &
                              combined_plot_data$Year %in% c(1/12, 6/12, 1, 3, 5, 10, 15, 20), ],
    aes(x = Year, y = Proportion, label = round(Proportion, 2)),
    vjust = -1, show.legend = FALSE
  ) +
  geom_text(
    data = combined_plot_data[combined_plot_data$Type == "Local" &
                              combined_plot_data$Year %in% c(3, 5, 10, 15, 20), ],
    aes(x = Year, y = Proportion, label = round(Proportion, 2)),
    vjust = -1, show.legend = FALSE
  ) +
  geom_text(
    data = combined_plot_data[combined_plot_data$Type == "Unknown" &
                              combined_plot_data$Year %in% c(3, 5, 10, 15, 20), ],
    aes(x = Year, y = Proportion, label = round(Proportion, 2)),
    vjust = 1.5, show.legend = FALSE
  ) +
  geom_text(
    data = combined_plot_data[combined_plot_data$Type == "Local&Distant" &
                              combined_plot_data$Year %in% c(3, 5, 10, 15, 20), ],
    aes(x = Year, y = Proportion, label = round(Proportion, 2)),
    vjust = 0, show.legend = FALSE
  )

```

## Kaplan Meier curves

The competing risk is going to be censored, we are overestimating the event of interest incidence by using Kaplan-Meier.

```{r}
library(survival)
library(survminer)
library(ggplot2)
```



```{r, fig.height=5, fig.weight=7}

# time_col   <- "Disease_Free_Survival_Months"
# status_col <- "Cmprsk_Status"
# 
# cat_vars <- names(df)[sapply(df, \(x) is.factor(x) || is.character(x))]
# cat_vars <- setdiff(cat_vars, c(time_col, status_col))
# 
# for (v in cat_vars) {
#   dat <- df[, c(time_col, status_col, v)]
# 
#   # make explicit columns so no symbols are looked up
#   dat$time  <- dat[[time_col]]
#   dat$event <- dat[[status_col]] == 1
#   dat$grp   <- droplevels(dat[[v]])
# 
#   fit <- survfit(Surv(time, event) ~ grp, data = dat)
# 
#   pl <- ggsurvplot(
#     fit, data = dat,
#     pval = TRUE, conf.int = TRUE, risk.table = TRUE,
#     ggtheme = theme_minimal(), title = paste("KM by", v), legend.title = v
#   )
# 
#   print(pl)
# }

```

```{r}
# for (v in cat_vars) {
#   dat <- df[, c(time_col, status_col, v)]
#   dat$time  <- dat[[time_col]]
#   dat$status <- dat[[status_col]]
#   dat$grp   <- droplevels(dat[[v]])
# 
#   ci <- cmprsk::cuminc(
#     ftime   = dat$time,
#     fstatus = dat$status,
#     group   = dat$grp
#   )
# 
#   plot(ci, wh = 1, main = paste0(v, " Cumulative incidence (cause 1)"), xlab = "Time (months)",
#      ylab = "Probability", lty=1:length(ci), color=1, lwd=par('lwd'),)
# 
# }
```


```{r  fig.height=5, fig.weight=7}
library(tidycmprsk)
library(ggplot2)
library(ggsurvfit)
library(cmprsk)
library(ggsurvfit)

time_col   <- "Time"
status_col <- "Outcome"

df2 <- df %>%
  mutate(
    !!status_col := factor(
      .data[[status_col]],
      levels = c(0, 1, 2), 
      labels = c("censor", "death from cancer", "death other causes")
    )
  ) %>%
  filter(is.finite(.data[[time_col]]), !is.na(.data[[status_col]]))

outcomes_present <- setdiff(levels(df2[[status_col]]), "censor")
cause_labels <- c("death from cancer", "death other causes")

grp_var <- "Sex"
df2[[grp_var]] <- factor(df2[[grp_var]])

f <- as.formula(paste0("Surv(", time_col, ", ", status_col, ") ~ ", grp_var))

ci_obj <- tidycmprsk::cuminc(f, data = df2)

pvals <- ci_obj$cmprsk$Tests[, "pv"] 
subtitle_txt <- paste(paste0(cause_labels, ": p=", signif(pvals, 3)), collapse = "   ")

p <- ggsurvfit::ggcuminc(ci_obj, outcome = c("death from cancer", "death other causes")) +
  add_confidence_interval() +
  add_risktable() +
  scale_ggsurvfit() +
  labs(
    title = paste("Cumulative incidence by", grp_var),
    subtitle = subtitle_txt,
    x = "Time (months)",
    y = "Cumulative incidence"
  )

print(p)

```

Let's perform this with all categoricals: 


```{r fig.height=5, fig.weight=7}
time_col   <- "Time"
status_col <- "Outcome"

df2 <- df %>%
  mutate(
    !!status_col := factor(
      .data[[status_col]],
      levels = c(0, 1, 2), 
      labels = c("censor", "death CRC/recurrence", "death other causes")
    )
  ) %>%
  filter(is.finite(.data[[time_col]]), !is.na(.data[[status_col]]))

outcomes_present <- setdiff(levels(df2[[status_col]]), "censor")
cause_labels <- c("death CRC/recurrence", "death other causes")

cat_vars <- names(df2)[vapply(df2, function(x) is.factor(x) || is.character(x) || is.logical(x), TRUE)]
cat_vars <- setdiff(cat_vars, c(time_col, status_col))


for (v in cat_vars) {

  f <- as.formula(paste0("Surv(", time_col, ", ", status_col, ") ~ ", v))
  
  ci_obj <- tidycmprsk::cuminc(f, data = df2)
  
  pvals <- ci_obj$cmprsk$Tests[, "pv"] 
  subtitle_txt <- paste(paste0(cause_labels, ": p=", signif(pvals, 3)), 
                        collapse = "   ")
  
  p <- ggcuminc(ci_obj, outcome = "death CRC/recurrence")+
    add_confidence_interval() +
    add_risktable() +
    scale_ggsurvfit() +
    labs(
      title = paste("Cumulative incidence by", v),
      subtitle = subtitle_txt,
      x = "Time (months)",
      y = "Cumulative incidence"
    )

  print(p)
}
```


## Missigness

```{r, fig.height=8}
library(naniar)

gg_miss_var(df, show_pct = TRUE)


```

## Correlation

```{r, fig.height=8}
library(reshape2)

num <- df %>%
  select(where(is.numeric))

corr <- stats::cor(num, use = "pairwise.complete.obs", method = "spearman")
melted <- melt(corr)

ggplot(melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, name = "Spearman ρ"
  ) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_fixed() +
  labs(title = "Correlation matrix (numeric covariates)", x = NULL, y = NULL)
```

SDMA and Beta Catenines might require selection. 

## Class imbalance

```{r, fig.height=8}
library(forcats)

df_long <- df %>%
  select(where(\(x) is.factor(x) || is.character(x))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  mutate(value = fct_explicit_na(as.factor(value), "NA"))

df_plot <- df_long %>%
  count(variable, value) %>%
  group_by(variable) %>%
  mutate(prop = n / sum(n))

ggplot(df_plot, aes(x = value, y = prop, fill = value)) +
  geom_col() +
  facet_wrap(~ variable, scales = "free_x") +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
            vjust = -0.3, size = 2.8) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Set3", na.value = "grey80") +
  labs(x = NULL, y = "Proportion", title = "Class imbalance across categorical variables") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

```


### Simple train/test split to test addition of covariates:


```{r}
library(recipes)
library(rsample)
library(survival)
library(cmprsk)
library(broom)
```

```{r}
set.seed(123)

rownames(df) <- df$TMA_Order

df <- df[, !(names(df) %in% c("TMA_Order"))]

time_col   <- "Time"
status_col <- "Outcome"

predictors <- setdiff(names(df), c(time_col, status_col))

predictors
```
```{r}
length(predictors)
```


Split by keeping the proportions of the outcome of interest. 

```{r}
# library(rsample)
# library(recipes)
# 
# set.seed(123)
# 
# split <- initial_split(df, prop = 0.8, strata = status_col)
# train <- training(split)
# test  <- testing(split)
# 
# # imputation to be applied to train
# 
# rx <- recipe(reformulate(predictors, response = time_col), data = train) %>%
#   # the outcome as time
#   update_role(all_of(time_col), new_role = "outcome") %>%
#   # the predictor is all the rest
#   update_role(all_of(setdiff(predictors, time_col)), new_role = "predictor") %>%
#   # median and mode
#   step_impute_median(all_numeric_predictors()) %>%
#   step_impute_mode(all_nominal_predictors()) %>%
#   step_zv(all_predictors()) %>%
#   # normalise the continuous
#   step_normalize(all_numeric_predictors(), -all_outcomes())
#   
# # imputation and scaling only from the training
# rx_prep <- prep(rx, training = train, verbose = FALSE)
# 
# X_train <- bake(rx_prep, new_data = train) %>% select(-all_of(time_col))
# X_test  <- bake(rx_prep, new_data = test)  %>% select(-all_of(time_col))
# 
# ftime_train   <- train[[time_col]]
# fstatus_train <- train[[status_col]]
# 
# ftime_test    <- test[[time_col]]
# fstatus_test  <- test[[status_col]]

```



```
library(mice)
library(rsample)
library(recipes)
library(dplyr)

set.seed(123)

# imputation
# restrict to predictors
pred_dat <- df[, predictors, drop = FALSE]

# let mice choose default methods
meth  <- make.method(pred_dat)
predM <- make.predictorMatrix(pred_dat)

imp <- mice(
  pred_dat,
  m      = 1, 
  method = meth,
  predictorMatrix = predM,
  maxit  = 100,
  seed   = 12,
  printFlag = FALSE
)

# completed predictors
pred_imp <- complete(imp, 1)

# full completed dataset
df_imp <- bind_cols(
  df[, setdiff(names(df), predictors), drop = FALSE],
  pred_imp
)

# split data
split <- initial_split(df_imp, prop = 0.8, strata = status_col)
train <- training(split)
test  <- testing(split)


# standarization
rx <- recipe(reformulate(predictors, response = time_col), data = train) %>%
  update_role(all_of(time_col),   new_role = "outcome") %>%
  update_role(all_of(predictors), new_role = "predictor") %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors(), -all_outcomes())

rx_prep <- prep(rx, training = train, verbose = FALSE)

X_train <- bake(rx_prep, new_data = train) %>% select(-all_of(time_col))
X_test  <- bake(rx_prep, new_data = test)  %>% select(-all_of(time_col))

ftime_train   <- train[[time_col]]
fstatus_train <- train[[status_col]]

ftime_test    <- test[[time_col]]
fstatus_test  <- test[[status_col]]

```

We are going to have an effect on the split and another effect from the imputation seed.

```{r}
library(recipes)
library(mice)

set.seed(43)

split <- initial_split(df, prop = 0.8, strata = status_col)
train <- training(split)
test  <- testing(split)

train_pred <- train[, predictors, drop = FALSE]
test_pred  <- test[,  predictors, drop = FALSE]

meth  <- make.method(train_pred)
predM <- make.predictorMatrix(train_pred)

imp_train <- mice(
  train_pred,
  m = 1,
  method = meth,
  predictorMatrix = predM,
  maxit = 100,
  seed = 12,
  printFlag = FALSE
)

train_pred_imp <- complete(imp_train, 1)

imp_test <- mice.mids(imp_train, newdata = test_pred)
test_pred_imp <- complete(imp_test, 1)

# adding time and outcome
train_imp <- bind_cols(train[, setdiff(names(train), predictors), 
                             drop = FALSE],
                       train_pred_imp)

test_imp  <- bind_cols(test[, setdiff(names(test), predictors), 
                            drop = FALSE],
                       test_pred_imp)

rx <- recipe(~ ., data = train_imp) %>%
  # update as outcome to avoid normalization of time
  update_role(all_of(time_col), new_role = "outcome") %>%
  update_role(all_of(status_col), new_role = "outcome") %>%
  update_role(all_of(predictors), new_role = "predictor") %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors(), -all_outcomes())

rx_prep <- prep(rx, training = train_imp, verbose = FALSE)

X_train <- bake(rx_prep, new_data = train_imp)
X_test  <- bake(rx_prep, new_data = test_imp)

ftime_train   <- train_imp[[time_col]]
fstatus_train <- train_imp[[status_col]]

ftime_test    <- test_imp[[time_col]]
fstatus_test  <- test_imp[[status_col]]
```


```{r}
imp_train$method
```



```{r}
library(ggmice)
#gmice(df, aes(IL17RA_Presence, Age)) +
#  geom_point()


ggmice(imp_train, aes(x = .imp, y = MMR_Status)) +
  geom_jitter(height = 0, width = 0.25) +
  #geom_boxplot(width = 0.5, size = 1, alpha = 0.75, outlier.shape = NA) +
  labs(x = "Imputation number")

ggmice(imp_test, aes(x = .imp, y = MMR_Status)) +
  geom_jitter(height = 0, width = 0.25) +
  #geom_boxplot(width = 0.5, size = 1, alpha = 0.75, outlier.shape = NA) +
  labs(x = "Imputation number")
```

```{r}
plot(imp_train) 
```


```{r}
train_imp
```



```{r}
densityplot(imp_train)
densityplot(imp_test)
```


Avoid collinearity:

```{r}
causes <- sort(setdiff(unique(fstatus_train), 0))

form_all <- as.formula(paste0("~ ", paste(colnames(X_train), collapse = " + ")))
MM <- model.matrix(form_all, data = X_train)[, -1, drop = FALSE]

# avoiding collinearity:
# but how has it been introduced? 
qrX <- qr(MM)
keep <- sort(qrX$pivot[seq_len(qrX$rank)])
predictors_train <- MM[, keep, drop = FALSE]

dif <- setdiff(colnames(MM), colnames(predictors_train))

dif
```

```{r}
# cor(train$Preoperative_CRP, train$CRP,
#   method = "spearman"
# )
# 
# cor(train$Preoperative_Lymphocytes, train$Lymphocytes,
#   method = "spearman"
# )

```

```{r}
#plot(log(train$Ki67_Image_Analysis), log(train$Ki67_Tumour))
```
```{r}
#plot(train$CD3FOXP3_Raw, train$CD3_FOXP3_Raw)
```
```{r}
#cor(train$CD3FOXP3_Raw, train$CD3_FOXP3_Raw, use="complete.obs")
```

```{r}
#plot(train$Preoperative_CRP, train$CRP)
```

```{r}
#plot(train$Preoperative_Lymphocytes, train$Lymphocytes)

```


```{r}
library(prodlim)
## fine and grey with cmprsk
# fit_fg <- lapply(causes, function(k) {
#   cmprsk::crr(
#     ftime   = ftime_train,
#     fstatus = fstatus_train,
#     cov1    = predictors_train,
#     failcode = k,
#     cencode  = 0
#   )
# })
# names(fit_fg) <- paste0("FG_cause_", causes)

## fine and grey with wrapper from riskregression
train_df <- as.data.frame(X_train)

predictors <- setdiff(names(train_df), c("Time", "Outcome"))

frm <- as.formula(paste0("Hist(Time, Outcome) ~ ", paste(predictors, collapse = " + ")))

fit_fg <- lapply(causes, function(k) {
  riskRegression::FGR(frm, data = train_df, cause = k)
})
names(fit_fg) <- paste0("FG_cause_", causes)

```

```{r}
length(predictors)
```


https://pmc.ncbi.nlm.nih.gov/articles/PMC5698744/

```{r, fig.height=8}

library(ggtext)

fg1 <- fit_fg$FG_cause_1$crrFit
s   <- summary(fg1)
tab <- as.data.frame(s$coef)
colnames(tab) <- make.names(colnames(tab), unique = TRUE)

#beta <- drop(fg1$coef)
#V    <- fg1$var
#se   <- sqrt(diag(V))
#z    <- beta / se
#pval <- 2 * pnorm(abs(z), lower.tail = FALSE)

HR <- data.frame(
  term       = rownames(tab),
  estimate   = tab[["exp.coef."]],
  conf.low   = exp(tab[["coef"]] - 1.96 * tab[["se.coef."]]),
  conf.high  = exp(tab[["coef"]] + 1.96 * tab[["se.coef."]]),
  p.value   = tab[["p.value"]],
  row.names  = NULL,
  check.names = FALSE
)

alpha <- 0.05
HR$sig <- ifelse(is.na(HR$p.value), "NA", ifelse(HR$p.value < alpha, "p<0.05", "ns"))
HR$sig <- factor(HR$sig, levels = c("p<0.05", "ns", "NA"))

HR <- HR %>%
  mutate(
    term_label = ifelse(sig == "p<0.05",
                        paste0("**", term, "**"),
                        term)
  )

ggplot(HR, aes(x = estimate, y = reorder(term_label, estimate), color = sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), na.rm = TRUE) +
  scale_x_log10() +
  scale_color_manual(
    values = c("p<0.05" = "#D55E00", "ns" = "#555555", "NA" = "#AAAAAA"),
    name = "Significance"
  ) +
  labs(x = "Hazard Ratio (log scale)", y = NULL, title = "Fine–Gray: Cause 1") +
  theme_bw(12) +
  theme(
    legend.position = "right",
    axis.text.y = element_markdown()
  )
```
```{r}
table(df$T_Stage, useNA="ifany") 
```


```{r}
table(train_df$T_Stage, useNA="ifany")
```

```{r}
#table(train_df$Tumour_Perforation, useNA="ifany")
```

```{r}
library(riskRegression)
library(prodlim)

# eval times
times <- quantile(ftime_test, probs = seq(0.1, 0.9, by = 0.05), na.rm = TRUE)
times <- sort(unique(as.numeric(times)))

X_test <- X_test[, colnames(X_train)]

test_df <- as.data.frame(X_test)

cif_fg1 <- predictRisk(fit_fg[["FG_cause_1"]], 
                       newdata = test_df, times = times)

#cif_fg1
```
```{r}
table(test_df$T_Stage, useNA="ifany")
```

```{r}
table(test_df$Tumour_Perforation, useNA="ifany")
```

```{r}
sc_fg <- Score(
  object      = list(FG = fit_fg[["FG_cause_1"]]),
  formula     = Hist(Time, Outcome) ~ 1,
  data        = test_df,
  times       = times,
  metrics     = "Brier",
  cause       = 1L,
  cens.model  = "km",
  summary     = "ibs",
  split.method = "none"
)
```

```{r}
sc_fg
```
```{r}

tail(sc_fg$Brier$score$IBS, 1)

```


### Additive performance of routine vs research variables.

```{r}
meta <- read.csv("data/GRI_preprocess-1-meta.csv", stringsAsFactors = FALSE)


routine_vars  <- meta %>% filter(Collection == "routine")
research_vars <- meta %>% filter(Collection == "research")
both_vars <- c("Time", "Outcome", "Lymph_Node_Ratio")


rou <- union(routine_vars$New.Data.Item, both_vars)

rou

```
```{r}
predictors

```


```{r}
res <- union(research_vars$New.Data.Item, both_vars)
res
```

```{r}
both <- intersect(c(union(rou, res)), names(train_df))

both
```

Using the same splits as before:

```{r}
library(prodlim)


train_df <- as.data.frame(X_train)

#train_df <- train_df[, !(names(train_df) %in% drop_vars)]

fit_models <- list()
for (nm in c("Routine", "Research", "Research/Routine")) {
  predictors <- switch(
    nm,
    "Routine"  = intersect(c(rou), names(train_df)),
    "Research" = intersect(c(res), names(train_df)),
    "Research/Routine"     = intersect(c(union(rou, res)), names(train_df))
  )
  
  preds <- setdiff(predictors, c("Time", "Outcome"))
  
  frm <- as.formula(paste0("Hist(Time, Outcome) ~ ", paste(preds, 
                                                          collapse = " + ")))
  fit_models[[nm]] <- riskRegression::FGR(frm, data = train_df, cause = 1)

  
}

# test prediction
test_df <- as.data.frame(X_test)
#test_df <- test_df[, !(names(test_df) %in% drop_vars)]

# evaluation
times_eval <- c(0, 6, 12, 36, 60, 120, 180) 

scores <- riskRegression::Score(
  object  = fit_models,
  formula = Hist(Time, Outcome) ~ 1,
  data    = test_df,
  times   = times_eval,
  metrics = "Brier",
  cause   = 1,
  summary = "ibs"
)

```

```{r}
df_brier <- scores$Brier$score

ggplot(df_brier, aes(x = factor(times), y = Brier, color = model, group = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  scale_y_continuous(name = "Brier score ", expand = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(name = "Time (months)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Brier score comparison across models")

```
```{r}
df_brier
```


```{r}
tidy_fg_hr <- function(fg_obj, model_name = "") {
  # accept FGR (wraps cmprsk::crr) or crr directly
  crr <- if (inherits(fg_obj, "FGR")) {
    if (!is.null(fg_obj$crrfit)) fg_obj$crrfit else fg_obj$crrFit
  } else fg_obj
  stopifnot(inherits(crr, "crr"))

  s   <- summary(crr)
  tab <- as.data.frame(s$coef)
  colnames(tab) <- make.names(colnames(tab), unique = TRUE)

  pcol <- intersect(c("p","p.value","Pr..>.z."), colnames(tab))
  pvec <- if (length(pcol)) tab[[pcol[1]]] else NA_real_

  out <- data.frame(
    term      = rownames(tab),
    estimate  = tab[["exp.coef."]],
    conf.low  = exp(tab[["coef"]] - 1.96 * tab[["se.coef."]]),
    conf.high = exp(tab[["coef"]] + 1.96 * tab[["se.coef."]]),
    p.value   = pvec,
    model     = model_name,
    row.names = NULL,
    check.names = FALSE
  )

  out$sig <- factor(
    ifelse(is.na(out$p.value), "NA",
           ifelse(out$p.value < 0.05, "p<0.05", "ns")),
    levels = c("p<0.05","ns","NA")
  )

  out$term_label <- ifelse(out$sig == "p<0.05",
                           paste0("**", out$term, "**"),
                           out$term)

  out
}

```

```{r}
models3 <- list(
  Routine  = fit_models[["Routine"]],
  Research = fit_models[["Research"]],
  `Research/Routine` = fit_models[["Research/Routine"]]
)
HR_all <- bind_rows(lapply(names(models3), function(nm) tidy_fg_hr(models3[[nm]], nm)))

HR_all <- HR_all %>%
  mutate(model = factor(model, levels = c("Routine", "Research", "Research/Routine"))) %>%
  group_by(model) %>%
  arrange(estimate, .by_group = TRUE) %>%
  mutate(term_key = paste(model, term_label, sep = "__")) %>% 
  ungroup()

HR_all$term_key <- factor(HR_all$term_key, levels = HR_all$term_key)
y_labs <- setNames(HR_all$term_label, HR_all$term_key)

ggplot(HR_all, aes(x = estimate, y = term_key, color = sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), na.rm = TRUE) +
  scale_x_log10() +
  scale_color_manual(
    values = c("p<0.05" = "#D55E00", "ns" = "#555555", "NA" = "#AAAAAA"),
    name   = "Significance"
  ) +
  scale_y_discrete(labels = y_labs) +
  labs(
    x     = "Hazard Ratio (log scale)",
    y     = NULL,
    #title = "Fine and Gray (Death of CRC/Recurrence): Hazard Ratios"
  ) +
  facet_wrap(~ model, scales = "free_y") +
  theme_bw(12) +
  theme(
    legend.position = "right",
    axis.text.y = element_markdown()
  )
```
#### try with other categorization: 


```{r}
meta <- read.csv("data/GRI_preprocess-1-meta.csv", stringsAsFactors = FALSE)


routine_vars  <- meta %>% filter(Collection == "routine")
research_vars <- meta %>% filter(Collection == "research")
ihc_vars <- meta %>% filter(Origin == "IHC")
non_ihc_vars <- meta %>% filter(Origin %in% c("Laboratory", 
                                              "MultiplexIF", 
                                              "RNAScope",
                                              "Pathology", 
                                              "Clinical", 
                                              "Genetics"))

both_vars <- c("Lymph_Node_Ratio", "Time", "Outcome")

rou <- union(routine_vars$New.Data.Item, both_vars)

ihc_rou <- union(routine_vars$New.Data.Item, both_vars)
ihc_rou <- union(ihc_rou, ihc_vars$New.Data.Item)

non_ihc_rou <- union(routine_vars$New.Data.Item, both_vars)
non_ihc_rou <- union(non_ihc_rou, non_ihc_vars$New.Data.Item)

```

```{r}
library(prodlim)


train_df <- as.data.frame(X_train)

#train_df <- train_df[, !(names(train_df) %in% drop_vars)]

fit_models <- list()
for (nm in c("Routine", "Routine/IHC", "Routine/non-IHC")) {
  predictors <- switch(
    nm,
    "Routine"      = intersect(c(rou), names(train_df)),
    "Routine/IHC"  = intersect(c(ihc_rou), names(train_df)),
    "Routine/non-IHC" = intersect(c(non_ihc_rou), names(train_df)), 
    
  )
  
  preds <- setdiff(predictors, c("Time", "Outcome"))
  
  frm <- as.formula(paste0("Hist(Time, Outcome) ~ ", paste(preds, 
                                                          collapse = " + ")))
  fit_models[[nm]] <- riskRegression::FGR(frm, data = train_df, cause = 1)

  
}

# test prediction
test_df <- as.data.frame(X_test)
#test_df <- test_df[, !(names(test_df) %in% drop_vars)]

# evaluation
times_eval <- c(0, 6, 12, 36, 60, 120, 180) 

scores <- riskRegression::Score(
  object  = fit_models,
  formula = Hist(Time, Outcome) ~ 1,
  data    = test_df,
  times   = times_eval,
  metrics = "Brier",
  cause   = 1,
  summary = "ibs"
)

```

```{r}
df_brier <- scores$Brier$score

ggplot(df_brier, aes(x = factor(times), y = Brier, color = model, group = model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  scale_y_continuous(name = "Brier score ", expand = expansion(mult = c(0, 0.05))) +
  scale_x_discrete(name = "Time (months)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Brier score comparison across models")

```
```{r}
df_brier
```




```{r}
tidy_fg_hr <- function(fg_obj, model_name = "") {
  # accept FGR (wraps cmprsk::crr) or crr directly
  crr <- if (inherits(fg_obj, "FGR")) {
    if (!is.null(fg_obj$crrfit)) fg_obj$crrfit else fg_obj$crrFit
  } else fg_obj
  stopifnot(inherits(crr, "crr"))

  s   <- summary(crr)
  tab <- as.data.frame(s$coef)
  colnames(tab) <- make.names(colnames(tab), unique = TRUE)

  pcol <- intersect(c("p","p.value","Pr..>.z."), colnames(tab))
  pvec <- if (length(pcol)) tab[[pcol[1]]] else NA_real_

  out <- data.frame(
    term      = rownames(tab),
    estimate  = tab[["exp.coef."]],
    conf.low  = exp(tab[["coef"]] - 1.96 * tab[["se.coef."]]),
    conf.high = exp(tab[["coef"]] + 1.96 * tab[["se.coef."]]),
    p.value   = pvec,
    model     = model_name,
    row.names = NULL,
    check.names = FALSE
  )

  out$sig <- factor(
    ifelse(is.na(out$p.value), "NA",
           ifelse(out$p.value < 0.05, "p<0.05", "ns")),
    levels = c("p<0.05","ns","NA")
  )

  out$term_label <- ifelse(out$sig == "p<0.05",
                           paste0("**", out$term, "**"),
                           out$term)

  out
}

```

```{r}
models3 <- list(
  Routine  = fit_models[["Routine"]],
  `Routine/IHC` = fit_models[["Routine/IHC"]],
  `Routine/non-IHC` = fit_models[["Routine/non-IHC"]]
)
HR_all <- bind_rows(lapply(names(models3), function(nm) tidy_fg_hr(models3[[nm]], nm)))

HR_all <- HR_all %>%
  mutate(model = factor(model, levels = c("Routine", "Routine/IHC", "Routine/non-IHC"))) %>%
  group_by(model) %>%
  arrange(estimate, .by_group = TRUE) %>%
  mutate(term_key = paste(model, term_label, sep = "__")) %>% 
  ungroup()

HR_all$term_key <- factor(HR_all$term_key, levels = HR_all$term_key)
y_labs <- setNames(HR_all$term_label, HR_all$term_key)

ggplot(HR_all, aes(x = estimate, y = term_key, color = sig)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), na.rm = TRUE) +
  scale_x_log10() +
  scale_color_manual(
    values = c("p<0.05" = "#D55E00", "ns" = "#555555", "NA" = "#AAAAAA"),
    name   = "Significance"
  ) +
  scale_y_discrete(labels = y_labs) +
  labs(
    x     = "Hazard Ratio (log scale)",
    y     = NULL,
    #title = "Fine and Gray (Death of CRC/Recurrence): Hazard Ratios"
  ) +
  facet_wrap(~ model, scales = "free_y") +
  theme_bw(12) +
  theme(
    legend.position = "right",
    axis.text.y = element_markdown()
  )
```